Accessing biological data with semantic web technologies

### 2013-12-26 article.tex
First code to create RDF from the BioMart data

\begin{table}
\caption{A list of ontology terms for the concepts in the BRCA1 SNP example. Ontology acronyms and the prefixes
used are explained in the text.}
\begin{center}
\begin{tabular}{l|l|l|l}
 \textbf{concept} & \textbf{ontology} & \textbf{term} & \textbf{URI} \\
\hline
 SNP & SIO & snp & sio:SIO\_010027 \\
\end{tabular}
\end{center}
\end{table}

Taking advantage of ...
SIO is the Semanticscience Integrated Ontology;
... Prefixes used in the term Uniform Resource Identifiers (URIs) include sio for \url{http://semanticscience.org/resource/} ...

@@

>>>>

\begin{table}
\caption{A list of ontology terms for the concepts in the BRCA1 SNP example. Ontology acronyms and the prefixes
used are explained in the text.}
\begin{center}
\begin{tabular}{l|l|l|l}
 \textbf{concept} & \textbf{ontology} & \textbf{term} & \textbf{URI} \\
\hline
 SNP & SIO & snp & sio:SIO\_010027 \\
 gene & SO & gene & obo:SO\_0000704 \\
 chromosome & GO & chromosome & obo:GO\_0005694 \\
 article &  BIBO & article & bibo:Article \\
 allele & SO & allele & obo:SO\_0001023 \\
 SNP identifier & TMO & snp identifier & tmo:TMO\_0161 \\
 PubMed identifier & CHEMINF & PubMed identifier & sio:CHEMINF\_000302\\
 chromosome number & TMO & chromosome nuber & tmo:TMO\_0157 \\
 chromosomal sequence position & TMO & chromosomal sequence position & tmo:TMO\_0122 \\
\end{tabular}CiTO
\end{center}
\end{table}

Taking advantage of existing ontologies. The choice is somewhat arbitrary and no ontological
analysis has been done, e.g. on specified domain and ranges of predicates and implied
properties of classes. The following ontologies are used:
SIO is the Semanticscience Integrated Ontology; TMO is the Translational Medicine Ontology;
SO is the Sequence Types and Features Ontology; GO is the Gene Ontology;
BIBO is the Bibliographic Ontology; CiTO is the Citation Typing Ontology;
... Prefixes used in the term Uniform Resource Identifiers (URIs) include
\textit{sio} for \url{http://semanticscience.org/resource/},
\textit{tmo} for \url{http://www.w3.org/2001/sw/hcls/ns/transmed/},
\textit{obo} for \url{http://purl.obolibrary.org/obo/},
\textit{bibo} for \url{http://purl.org/ontology/bibo/},
\textit{cito} for \url{} ...

Using these ontology terms we can define the translations of the BioMart-extracted
data into RDF:

\begin{verbatim}
snpStore = new.rdf(ontology=FALSE)
rdfTypePred = "http://www.w3.org/1999/02/22-rdf-syntax-ns#type"
snpClass = "http://semanticscience.org/resource/SIO_010027"
createEntry <- function(row) {
  snpID = row[1]
  snpSubject = paste("http://example.org/snp/", snpID, sep="")
  add.triple(snpStore,
    subject=snpSubject, predicate=rdfTypePred, object=snpClass
  )
}
\end{verbatim}

We can then just iterate over the actual data with:

\begin{verbatim}
apply(data, MARGIN=1, FUN=createEntry(row) )
\end{verbatim}

@@

%\subsection{Example: BioPortal}
%
%In the BRCA1 SNP example, we found ontology terms for concepts using BioPortal. Now, at the
%time of writing BioPortal has an experimental SPARQL end point which allows us to find terms
%in this way too. For example, to find terms related to SNPs, we can run this code in R:
%
%\begin{verbatim}
%findSNPterms = paste(
%  "PREFIX rdfs: <http://www.w3.org/2000/01/rdf-schema#>",
%  "SELECT DISTINCT ?term ?label WHERE {",
%  " ?term rdfs:label ?label .",
%  "FILTER (CONTAINS (LCASE(str(?label)), \"snp\"))",
%  "}"
%)
%sparql.remote(
%  endpoint="http://sparql.bioontology.org/sparql",
%  sparql=findSNPterms, jena=FALSE
%)
%\end{verbatim}
